---
title: "APEX active layer analysis"
author: "Manies"
date: '2022-03-23'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Needed libraries:
```{r message=FALSE}
library(lubridate)
library(ggplot2)
library(dplyr)   #left_join

```

## Importation of data files
The data included in this analysis are from various sources.
All data are originally from Excel files where a tab was saved in csv format.  
Once imported each file needed their time stamp converted into strptime.

```{r}
guelph <- read.csv("raw_data/Seasonal Thaw Depth Data Dielman.csv",header=TRUE)
usgs1 <- read.csv("raw_data/FINAL End of Season usgs.csv",header=TRUE)
crell <- read.csv("raw_data/douglas APEX ERT lines thaw depths October 2017.csv", header=TRUE)
```
Add a column to each dataset so we know where the data is from.
```{r}
usgs1 <- cbind(usgs1, source="USGS")
guelph <- cbind(guelph, source="Guelph")
crell <- cbind(crell, source="CRELL")
```

## Basic review of each dataset by plotting data
### Guelph data from Katherine Dielman:
Measurements with permafrost are called "Permafrost Plateau" under Site.<br /> 
Names changed throughout the study time. Here old names are converted to current names.
```{r message=FALSE}
lookup <- read.csv("raw_data/guelph_lookup.csv",header=TRUE)
lookup$Plot <- as.character(lookup$Plot)
guelph <- left_join(guelph, lookup, by = c("Treatment" , "Plot"))
guelph$new_plot2 <- ifelse(is.na(guelph$new_plot),guelph$Plot,guelph$new_plot)

guelph.gamma <- subset(guelph, Site=="Permafrost Plateau ")
guelph.gamma$Date <- strptime(guelph.gamma$Date, format="%m/%d/%Y", tz="America/Anchorage")
guelph.gamma$month <- month(guelph.gamma$Date)
guelph.gamma.AugOn <- subset(guelph.gamma, month=="8" | month=="9" | month=="10")
aggregate(Ice_Depth_cm ~ Year, guelph.gamma.AugOn, mean)
g.gamma.AugOn.plot <- ggplot() +
  geom_point(data=guelph.gamma.AugOn, aes(x=as.POSIXct(Date), y=Ice_Depth_cm, color=factor(new_plot2)))
print(g.gamma.AugOn.plot)
```

QUESTION: How do these sites relate to each other spatially?

### USGS
Measurements with permafrost are labeled "forest" under Treatment
Need to convert the date column from a character to a date
```{r}
usgs.gamma <- subset(usgs1, usgs1$Treatment == "forest")
usgs.gamma$date <- strptime(usgs.gamma$date, format="%d-%b-%y", tz="America/Anchorage")
usgs.gamma.AugOn <- subset(usgs.gamma, format(usgs.gamma$date,"%b")=="Aug" | format(usgs.gamma$date,"%b")=="Sept")
aggregate(AL_cm ~ year, usgs.gamma.AugOn, mean)
usgs.gamma.Augon.plot <- ggplot(usgs.gamma.AugOn, aes(x=as.POSIXct(date), y=AL_cm)) +
  geom_point(aes(color=factor(Chamber)))
print(usgs.gamma.Augon.plot)
```
USGS data only until 2015. We must have more active layer measurements (associate with the BZPC cores) that we can include.

### CRELL data from Tom Douglas
Type of site was assigned by Kristen Manies based on the descriptions. 
Anything that said it had live spruce was assumed to have permafrost. 
Anything labeled thermokarst, noted to have standing water, or dead spruce was assumed to be thawed.
For the few rows without descriptions the determination was based on adjacent rows and/or the depth measurements.
```{r}
crell.gamma <- subset(crell, Site=="Pfrost")
#extract date from timestamp
crell.gamma$date <-substr(crell.gamma$DateTime, 1, 10)
crell.gamma$transect <- substr(crell.gamma$Name, 4,5)
crell.gamma$year <- year(crell.gamma$date)
aggregate(Depth ~ transect, crell.gamma, mean)
crell.gamma$date <- as.POSIXct(crell.gamma$date, "%Y-%m-%d", tz="America/Anchorage")
crell.gamma.plot <- ggplot(crell.gamma, aes(x=as.POSIXct(date), y=Depth)) +
  geom_point(aes(color=factor(transect)))
print(crell.gamma.plot)

```

## Combining all datasets
Combine gamma data from all three data sets and plot over time.
Columns in new dataframe will be: site (aka within gamma location), Date (%Y-%m-%d format), and active layer depth

```{r}
cg.sub <- data.frame(crell.gamma$transect, crell.gamma$date, crell.gamma$Depth, crell.gamma$source)
gg.sub <- data.frame(guelph.gamma$Treatment, guelph.gamma$Date, guelph.gamma$Ice_Depth_cm, guelph.gamma$source)
usgs.sub <- data.frame(usgs.gamma$Chamber, usgs.gamma$date, usgs.gamma$AL_cm, usgs.gamma$source)
same.names <- function(df) {
  names(df)[1] <- "location"
  names(df)[2] <- "date"
  names(df)[3] <- "depth"
  names(df)[4] <- "source"
  return(df)
}
cg.sub <-same.names(cg.sub)
gg.sub <-same.names(gg.sub)
usgs.sub <- same.names(usgs.sub)
all.gamma <- rbind(cg.sub, gg.sub, usgs.sub)
```
Plot the data
```{r}
all.plot <- ggplot(data=all.gamma, aes(x=as.POSIXct(date), y=depth)) +
  geom_point(aes(color=factor(source))) +
  ggtitle("Entire growing season")
print(all.plot)

august.on <- subset(all.gamma, month(date)>=8)
aug.plot <- ggplot(data=august.on, aes(x=as.POSIXct(date), y=depth)) +
  geom_point(aes(color=factor(source))) +
  ggtitle("August and later")
print(aug.plot)

```


